---
title: "ESM 244 Lab 1 Rathbone"
author: "Vanessa Rathbone"
date: "1/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Attach packages
library(tidyverse)
library(here)
library(sf)
library(tmap)

```

### Read in SF trees data 
```{r}

#read in data
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))

```

### Basic wrangling reminders 

Refresh some skills for data wrangling and summary statistics using functions in the 'dplyr' package.

Find the top 5 highest observations of trees by legal_status, do some wrangling, make a graph.
```{r}

#wrangle and clean up data
#count is 1) recognising groups, 2) replaces N function, 3) and summarizes function

top_5_status <- sf_trees %>% 
  count(legal_status) %>% #could add another group levels like count(legal_status, species)
  drop_na(legal_status) %>% #drop any NA values, in sf_trees it's group 10
  rename(tree_count = n) %>% 
  relocate(tree_count) %>%   #move columns around in df, similar to select function but simpler for multiple columns
  slice_max(tree_count, n = 5) #like head function but interactive if more values are added in the future, finds top 5 in this case
```

Make a graph of those top 5 observations by legal status. 
```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count))+
  geom_col() +
  labs(x = "Legal Status", y = "Tree Count") +
  coord_flip() + 
  theme_minimal()

```

### A few more data wrangling refresher examples

Only want to keep observations (rows) for Blackwood Acacia trees. 
```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% #look within a specified variable to only keep within species where blackwood acacia is detected 
  select(legal_status, date, latitude, longitude)

#preview the new df
ggplot(data = blackwood_acacia, aes(x = longitude, y = latitude)) +
  geom_point()

```

### tidyr::separate() and unite() functions

Useful for combining and separating columns

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep = "::")

```

Example: tidyr::unite()
```{r}
sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id:legal_status, sep = "_cool!_") #combining variables with a seperator such as _cool!_ in this example

```

